---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First we are loading the necessary libraries

```{r, warning=FALSE, message = FALSE}

library(readr)
library(tidyverse)
library(Hmisc)

```
Now we load the data specifying the date format
```{r, warning=FALSE, message = FALSE}

activity <- read_csv("activity.zip", col_types = cols(date = col_date(format = "%Y-%m-%d")))

```
## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day

```{r, warning=FALSE, message = FALSE}

steps_per_day <- activity %>% group_by(date) %>% summarise(sum(steps)) %>%  
                            rename("sum_of_steps" = "sum(steps)")

qplot(steps_per_day$sum_of_steps, 
      geom = "histogram", 
      main = "Steps taken each day", 
      xlab = "Steps", 
      ylab = "Frequency", 
      fill = I("blue"), 
      col = I("red"), 
      alpha = I(.2))
```

2. Calculate and report the **mean** and **median** total number of steps taken per day

```{r, warning=FALSE, message = FALSE}


mean_of_steps <- round(mean(steps_per_day$sum_of_steps, na.rm = TRUE))
mean_of_steps

median_of_steps <- round(median(steps_per_day$sum_of_steps, na.rm = TRUE))
median_of_steps

```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, warning=FALSE, message = FALSE}

avg_steps <- aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)

ggplot(data = avg_steps, aes(x = interval, y = steps)) +
        geom_line() +
        xlab("Intervals") +
        ylab("Average number of steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, warning=FALSE, message = FALSE}
max_steps <- avg_steps %>% arrange(desc(steps)) %>% head(1)
max_steps
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, warning=FALSE, message = FALSE}
total_NAs <- sum(is.na(activity))
total_NAs
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

There is a hint in the question: **impute** is a function and we can use **mean** with it as suggested.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, warning=FALSE, message = FALSE}
activity_no_NAs <- activity
activity_no_NAs$steps <- impute(activity_no_NAs$steps, fun = mean)
head(activity_no_NAs)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the **mean** and **median** total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r, warning=FALSE, message = FALSE}
steps_no_NAs <- activity_no_NAs %>% group_by(date) %>% summarise(sum(steps)) %>% 
                                    rename("sum_of_steps" = "sum(steps)")

qplot(steps_no_NAs$sum_of_steps, 
      geom = "histogram", 
      main = "Steps taken each day (with imputed missing values)", 
      xlab = "Steps", 
      ylab = "Frequency", 
      fill = I("blue"), 
      col = I("red"), 
      alpha = I(.2))

mean_of_steps_no_NAs <- round(mean(steps_no_NAs$sum_of_steps, na.rm = TRUE))
mean_of_steps_no_NAs

median_of_steps_no_NAs <- round(median(steps_no_NAs$sum_of_steps, na.rm = TRUE))
median_of_steps_no_NAs
```

The **mean** hasn't changed while the **median** has changed slightly. Imputing  NAs has had little effect on the data.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r, warning=FALSE, message = FALSE}
activity_no_NAs$day_type <-  ifelse(as.POSIXlt(activity_no_NAs$date)$wday %in% c(0,6), 'weekend', 'weekday')
head(activity_no_NAs)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).


```{r, warning=FALSE, message = FALSE}
avg_steps_no_NAs <- aggregate(steps ~ interval + day_type, data=activity_no_NAs, mean)

ggplot(avg_steps_no_NAs, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(day_type ~ .) +
    xlab("Interval") + 
    ylab("Number of steps")
```
